Model Finding for Recursive Functions in SMT

### 2015-04-29 tex/main.tex
Rearranging, add intro text to main section, more additions.

@@

\subsection{Model Finding for Quantified Formulas in SMT}

z3, cvc4

\section{Approach}

\subsection{Preliminaries}
\label{sec:prelim}

\subsection{Encoding Well-Defined Assumption for Recursive Functions}
\label{sec:encoding}

In the remainder of the section, we fix a signature $\Sigma$.

Say that $\Sigma$-formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$,
and the definitions in $\sfundefs{\Sigma}$ are logically consistent \ajr{formalize}.
In this section, we present a method which constructs an extended signature $\Sigma'$,
and translates $\varphi$ to equisatisfiable $\Sigma'$-formula $\psi$ on which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
The intuition behind this translation is use an uninterpreted type to abstract the set of tuples that each recursive function $f$ is invoked on,
and restrict the quantification of the function definition for $f$ to a variable of this type.

@@

Given the extended signature, we run the procedure $\conv$ in Figure~\ref{fig:encoding} on $\varphi$.
This calls the subprocedure $\conv_0$, which takes two arguments: the term $t$ to translate, and a \emph{polarity} $p$ (either $\ptrue$, $\pfalse$, or $\pnone$).
It returns a pair of the form $( r, D )$, where $r$ is a term of type $\tau$, and $D$ is a set of formulas.
%The role of $D$ is to ensure that the (restricted) function definition for $f$ includes certain tuples in its domain, which we explain more in the following.
At a high level, the translation modifies $\varphi$ in two ways.
First, it restricts the quantification on function definitions for $f$ to the corresponding uninterpreted sort $\fargsort{f}$.
Second, it augments $\varphi$ with additional constraints of the form $\exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t )$
which ensure that this restriction for the defintion $f$ covers all relevant tuples of terms,
namely the ones for which an $f$-application exists and is relevant to the satisfiability of the current branch of $\varphi$.

@@

We demonstrate this translation with a simple example.

\begin{example}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$ and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{equation*}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation*}
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universal quantification in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the sort $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of sort $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{equation*}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{equation*}
The universal quantification in the resulting formula is over an uninterpreted sort $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
$\square$
\end{example}

It remains to show that our translation produces an equisatisfiable formulae.
Namely,
we claim that if $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$,
function definitions in $\sfundefs{\Sigma}$ are logically consistent \ajr{formalize},
then $\conv(\varphi)$ is equisatisfiable to $\varphi$.

\begin{lemma}
Let $\varphi$ be a $\Sigma$-formula that is well-defined with respect to functions $\sfundefs{\Sigma}$.
\ajr{todo}
\end{lemma}

@@

For benchmarks, we gathered a set of 245 benchmarks from two sources, which we will refer to as {\bf isa} and {\bf leon}.
The first ({\bf isa}) consists of the 79 benchmarks from the Isaplanner benchmark suite~\cite{DBLP:conf/itp/JohanssonDB10} that did not contain higher-order functions.
The benchmarks have been used recently as challenge problems for a variety of inductive theorem provers.
These benchmarks heavily involve recursive functions, and are limited to the combined theory of uninterpreted functions and inductive datatypes.
The second ({\bf leon}) consists of 166 benchmarks from the Leon repository~\footnote{Available at \url{https://github.com/epfl-lara/leon}.},
which were constructed from verification conditions from simple Scala programs.
These benchmarks also heavily involve recursive functions over inductive datatypes, 
and also cover a wide variety of theories, including bitvectors, arrays, and linear and non-linear arithmetic.
Each of the 245 benchmarks are of the form described in Formula (~\ref{eq:wdf}) from Section~\ref{sec:prelim},
namely they each are a conjoined list of function definitions followed by a (negated) conjecture.
A majority of these benchmarks are unsatisfiable, although a handful of the benchmarks in the {\bf leon} set are satisfiable.

@@

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008}, 
which runs both heuristic methods for quantifier instantiation~\cite{DBLP:conf/cade/MouraB07},
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered three configurations of \cvc~\cite{barrett-et-al-2011} which we will refer to as {\bf cvc4}, {\bf cvc4+f} and {\bf cvc4+i}.
The default configuration {\bf cvc4} runs heuristic quantifier instantiation, 
but does not include techniques for finding models.
The configuration {\bf cvc4+f} runs a finite model finding procedure~\cite{ReyEtAl-1-RR-13, reynolds-et-al-2013}.
The last configuration {\bf cvc4+i} incorporates techniques for automating inductive reasoning in SMT~\cite{reynolds-kuncak-2015}.
%All configurations of \cvc incorporate techniques for conflict-based instantiation~\cite{} ..?

@@

The results are shown in Figure~\ref{fig:results}.
The results are divided into benchmarks before and after the translation $\conv$.
First, we note that in no cases did a configuration answer differently across the two encodings.
In other words, there were no cases in which \ziii or \cvc answered e.g. ``unsatisfiable" on a benchmark $\varphi$ in {\bf isa} 
and ``satisfiable" on its corresponding benchmark $\conv( \varphi )$ in {\bf isa+w}.
This provides confirming evidence that the translation $\conv$ is \emph{sound},
as we did not find a case where $\conv( \varphi )$ was unsatisfiable when $\varphi$ was satisfiable.
It also provides confirming evidence that the translation is \emph{complete} for this set of benchmarks,
as we did not find a case where $\conv( \varphi )$ was satisfiable when $\varphi$ was unsatisfiable.
This is to be expected, since the functions in our benchmark sets are intended to be logically consistent.

@@

Now, examining the second encoding (the latter nine columns), 
we see that using the translation $\conv$ drastically improves the ability of configurations for answering satisfiable.
In total, {\bf cvc4+f} finds 331 of the 885 benchmarks to be satisfiable, including 9 benchmarks in the original {\bf leon} benchmark set.
The performance of {\bf z3} for counterexamples also improves drastically, as it solves 112 satisfiable,
including 10 that were unsolved by {\bf cvc4+f}.
Moreover, the translation $\conv$ helps \ziii for unsatisfiable responses as well,
as we see \ziii solves a total of 209 whereas it solves only 187 before the translation.
This number is 2 fewer than {\bf cvc4}, which solves 211 unsatisfiable.
Thus, in most cases the translation $\conv$ does not seriously degrade performance for unsatisfiable benchmark,
and in some cases may actually aid the solver for determining unsatisfiability
\footnote{The exception here is {\bf cvc4+i}, which makes use of type information of quantified formulas to make inductive arugments.}.

>>>>

\section{Preliminaries}
\label{sec:prelim}

@@

\section{Model Finding for Recursive Functions in SMT}

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} 
have focused on finding models for quantified formulas in SMT.
A instatiation-based procedure is shown in~\cite{GeDeM-CAV-09} to be complete for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all occurrences of variables occur as direct subchildren of uninterpreted functions.
For instance, the formula $\forall x\typ{\sortint}. f( x ) \teq g( x ) + 5$ 
is in the essentially uninterpreted fragment since all occurrences of $x$ occur beneath uninterpreted functions $f$ and $g$.
Other syntactic criteria are identified in this paper that extend this fragment slightly, including cases when variables occur as arguments of arithmetic predicates.
The techniques in~\cite{ReyEtAl-1-RR-13} can be used for finding finite models for quantified formulas over uninterpreted types as well as types having a 
fixed finite interpretation (such as fixed-width bitvectors, datatype enumerations, etc.).
For instance, 
finite model finding techniques can find a model for
the formula $\forall x\typ{\tau} y\typ{\tau} z\typ{\tau}. x \teq y \vee y \teq z \vee x \teq z$
where the uninterpreted type $\tau$ is interpreted as a set of (at most) two elements.

Unfortunately, these fragments do not coincide with the vast majority of quantified formulas that correspond to recursive function definitions.
In particular, the fragments in~\cite{GeDeM-CAV-09} are not applicable when the argument of a recursive function is used in various ways (e.g. within an arithmetic expression), 
while the fragment in~\cite{ReyEtAl-1-RR-13} is not applicable for functions over infinite domains such as the integers.
So, for instance, the quantified formula $\forall x\typ{\sortint}. \lite( p \leq 0, p( x ) \teq 0, p( x ) \teq 2 * p( x - 1 ) )$ 
is not in the essentially uninterpreted fragment (since $x$ occurs as a child of $-$),
nor does finite model finding apply (since $x$ is of type $\sortint$).

Here, we advocate an approach that translates formulas involving function definitions (such as the one mentioned in the previous paragraph)
into equisatisfiable formulas where the techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.

\subsection{A Translation for Function Definitions}
\label{sec:encoding}

In the remainder of the section, we fix a signature $\Sigma$.
Say that $\Sigma$-formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$, 
and the definitions in $\sfundefs{\Sigma}$ are logically consistent.
We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
In particular, this section presents a method for constructing an extended signature $\extendsig( \Sigma )$,
and a $\extendsig( \Sigma )$-formula $\psi$ such that $\varphi$ and $\psi$ are equisatisfiable.
The intuition behind this translation is use an uninterpreted type to abstract the set of \emph{relevant} tuples for each recursive function $f$,
and restrict the quantification of the function definition for $f$ to a variable of this type.
Informally, the relevant tuples $\vec t$ of a function $f$ are the ones for which the interpretation of $f( \vec t )$ is relevant to the satisfiability of $\varphi$.

@@

Given the extended signature, we run the procedure $\conv$ in Figure~\ref{fig:encoding} on $\varphi$.
This calls the subprocedure $\conv_0$, which takes two arguments: the term $t$ to translate, and a \emph{polarity} $p$ (either $\ptrue$, $\pfalse$, or $\pnone$).
It returns a pair of the form $( r, D )$, where $r$ is a term of type $\tau$, and $D$ is a set of formulas.
%The role of $D$ is to ensure that the (restricted) function definition for $f$ includes certain tuples in its domain, which we explain more in the following.
At a high level, the translation modifies $\varphi$ in two ways.
First, it restricts the quantification on function definitions for $f$ to the corresponding uninterpreted type $\fargsort{f}$.
Second, it augments $\varphi$ with additional constraints of the form $\exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t )$
which ensure that this restriction for the defintion $f$ covers all relevant tuples of terms,
namely the ones for which an $f$-application exists and is relevant to the satisfiability of the current branch of $\varphi$.

@@

We demonstrate this translation with an example.

\begin{example}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$ and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{eqnarray} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{eqnarray} 
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the type $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of type $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
The universal quantification in the resulting formula is over an uninterpreted type $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
We observe that both \cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}),
and both produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.
$\square$
\end{example}

It is important to note that the translation $\conv$ does not preserve the models of $\varphi$.
One model $\M$ for formula~(\ref{eq:ex-after}) in the previous example interprets 
$\fargsort{s}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{s}$ as a finite map mapping containing $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$c$ as $14$, 
and $s$ as the (almost constant) function:
\begin{eqnarray}
\lambda x\typ{Int}. \lite( x \teq 0, 0, \lite( x \teq 1, 1, \lite( x \teq 2, 3, \ldots, \lite( x \teq 13, 91, 105 ) \ldots )
\end{eqnarray}
In other words, $s$ is a function mapping $x$ to the sum of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$,
and $105$ otherwise.
Notice that $\M$ is \emph{not} a model for the original formula~(\ref{eq:ex-before}),
since $\M$ interprets e.g. $s( -1 )$ and $s( 15 )$ as $105$.
However, under the assumption that function definitions in $\sfundefs{\Sigma}$ are logically consistent, 
we claim that $\conv(\varphi)$ is equisatisfiable to $\varphi$.
The intuition is that the interpretation of a term such as $s( -1 )$ is not relevant to the satisfiability of~(\ref{eq:ex-before}).
\ajr{add invariants, proofs}

\begin{theorem}
If $\varphi$ be a $\Sigma$-formula that is well-defined with respect to functions $\sfundefs{\Sigma}$,
and function definitions in $\sfundefs{\Sigma}$ are logically consistent,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}

@@

For benchmarks, we gathered a set of 245 benchmarks from two sources, which we will refer to as {\bf isa} and {\bf leon}.
The first ({\bf isa}) consists of the 79 benchmarks from the Isaplanner benchmark suite~\cite{DBLP:conf/itp/JohanssonDB10} that did not contain higher-order functions.
The benchmarks have been used recently as challenge problems for a variety of inductive theorem provers.
These benchmarks heavily involve recursive functions, and are limited to the combined theory of uninterpreted functions and inductive datatypes.
The second ({\bf leon}) consists of 166 benchmarks from the Leon repository~\footnote{Available at \url{https://github.com/epfl-lara/leon}.},
which were constructed from verification conditions from simple Scala programs.
These benchmarks also heavily involve recursive functions over inductive datatypes, 
and also cover a wide variety of theories, including bitvectors, arrays, and linear and non-linear arithmetic.
Each of the 245 benchmarks are of the form described in formula~(\ref{eq:wdf}) from Section~\ref{sec:prelim},
in particular they each are a conjoined list of function definitions followed by a (negated) conjecture.
A majority of these benchmarks are unsatisfiable, although a handful of the benchmarks in the {\bf leon} set are satisfiable.

@@

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008}, 
which runs both heuristic methods for quantifier instantiation~\cite{DBLP:conf/cade/MouraB07},
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered three configurations of \cvc~\cite{barrett-et-al-2011} which we will refer to as {\bf cvc4}, {\bf cvc4+f} and {\bf cvc4+i}.
The default configuration {\bf cvc4} runs heuristic quantifier instantiation, 
but does not include techniques for finding models.
The configuration {\bf cvc4+f} runs heuristic instantiation and the finite model finding procedure described in~\cite{ReyEtAl-1-RR-13, reynolds-et-al-2013}.
The last configuration {\bf cvc4+i} incorporates techniques for automating inductive reasoning in SMT~\cite{reynolds-kuncak-2015}.
%All configurations of \cvc incorporate techniques for conflict-based instantiation~\cite{} ..?

@@

The results are shown in Figure~\ref{fig:results}.
The results are divided into benchmarks before and after the translation $\conv$.
First, we note that in no cases did a configuration answer differently across the two encodings.
In other words, there were no cases in which \ziii or \cvc answered e.g. ``unsatisfiable" on a benchmark $\varphi$
and ``satisfiable" on its corresponding benchmark $\conv( \varphi )$, or vice versa.
This provides confirming evidence that the translation $\conv$ is \emph{sound},
as we did not find a case where $\conv( \varphi )$ was unsatisfiable when $\varphi$ was satisfiable.
It also provides confirming evidence that the translation is \emph{complete} for this set of benchmarks,
as we did not find a case where $\conv( \varphi )$ was satisfiable when $\varphi$ was unsatisfiable.
This is to be expected, since the functions in our benchmark sets are intended to be logically consistent.

@@

Now, examining the second encoding (the latter nine columns), 
we see that using the translation $\conv$ drastically improves the ability of configurations for answering satisfiable.
In total, {\bf cvc4+f} finds 331 of the 885 benchmarks to be satisfiable, including 9 benchmarks in the original {\bf leon} benchmark set.
The performance of {\bf z3} for counterexamples also improves drastically, as it solves 112 satisfiable,
including 10 that were unsolved by {\bf cvc4+f}.
Moreover, the translation $\conv$ helps \ziii for unsatisfiable responses as well,
as we see \ziii solves a total of 209 whereas it solves only 187 before the translation.
This number is 2 fewer than {\bf cvc4}, which solves 211 unsatisfiable.
Thus, in most cases the translation $\conv$ does not seriously degrade performance for unsatisfiable benchmark,
and in some cases may actually aid the solver for determining unsatisfiability
\footnote{The exception here is {\bf cvc4+i}, which makes use of type information associated with quantified formulas for inductive arguments.}.

We conclude from this evaluation that the translation $\conv$ enables SMT solvers to find counterexamples 
for conjectures involving recursive functions whose definitions are logically consistent.
Both {\bf z3} and {\bf cvc4+f} were able to find a large number of counterexamples after the translation,
with the latter configuration finding significantly more instances (331 to 112).

\section{Front-End Support for Recursive Function Definitions}
\label{sec:front-end}

We have seen a satisfiability-preserving translation $\conv$ that significantly improves the effectiveness of known SMT techniques
for finding counterexamples for formulas involving recursive function definitions.
For convienience, the translation $\conv$ has been implemented in the latest development version of \cvc~\cite{ReyEtAl-1-RR-13}.
